---
title: "Client Report: Central England Seasonal Temperature Change"
author: "Team 3: Yueran Wang 2201370"
date: "2026-04-01"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    pandoc_args:
      - "--shift-heading-level-by=-1"
header-includes:
  - \usepackage{float}
  - \usepackage{placeins}
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath(".."))

cache_root <- normalizePath(file.path("cache"), winslash = "/", mustWork = FALSE)
dir.create(cache_root, recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(cache_root, "knitr"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(cache_root, "figures"), recursive = TRUE, showWarnings = FALSE)

src_files <- list.files("src", pattern = "\\.R$", full.names = TRUE)
src_stamp <- if (length(src_files) == 0) {
  "no_src_files"
} else {
  paste(file.info(src_files)$mtime, collapse = "|")
}

data_files <- list.files("processed", pattern = "\\.csv$", full.names = TRUE)
data_stamp <- if (length(data_files) == 0) {
  "no_processed_csv"
} else {
  paste(file.info(data_files)$mtime, collapse = "|")
}

force_recompute <- FALSE

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.pos = "H",
  fig.align = "center",
  out.width = "88%",
  cache = TRUE,
  cache.path = file.path(cache_root, "knitr", "client_report", ""),
  fig.path = file.path("cache", "figures", "client_report-"),
  autodep = TRUE,
  cache.lazy = FALSE,
  cache.comments = FALSE,
  cache.extra = list(src_stamp, data_stamp, force_recompute, R.version.string)
)
```

```{r eda-sources, include=FALSE, cache=FALSE}
source(file.path("src", "eda_utils.R"))
source(file.path("src", "plot_eda_annotated_series.R"))
source(file.path("src", "plot_eda_warming_rate_curve.R"))
```

```{r model-sources, include=FALSE, cache=FALSE}
source(file.path("src", "table_utils.R"))
source(file.path("src", "modeling_gls_ar1.R"))
source(file.path("src", "model_plot_baseline_shift_gls_ar1.R"))
source(file.path("src", "model_plot_long_run_trend_gls_ar1.R"))
source(file.path("src", "model_thresholds.R"))
source(file.path("src", "table_thresholds.R"))
source(file.path("src", "table_monitoring_framework.R"))
```

```{r model-fit, include=FALSE, cache=TRUE}
model_df <- prepare_model_data(
  year_min         = 1879,
  year_max         = 2024,
  historical_start = 1961,
  historical_end   = 1990,
  recent_start     = 1991,
  recent_end       = 2020
)

model_results <- run_gls_models(model_df)
```

```{r monitoring-thresholds-compute, include=FALSE, cache=TRUE}
monitoring_thresholds    <- compute_monitoring_thresholds(recent_start = 1991, recent_end = 2020)
monitoring_thresholds_10 <- compute_monitoring_thresholds(recent_start = 2015, recent_end = 2024)
```

\newpage

## Executive summary

### The client question

The brief asked for an evidence-based view of how Central England temperatures have changed across the four seasons, and what those changes mean for internal planning. Three practical questions drove the work, covering whether recent decades are measurably warmer than an earlier generation, which seasons have changed the most, and what monitoring thresholds to use in seasonal briefings. This report addresses all three.

### Headline findings

- Spring temperatures averaged roughly 1.1 degrees Celsius higher in the 1991 to 2020 period than in 1961 to 1990, the largest shift of the four seasons. Summer and autumn also show clear increases. Winter points in the same direction with a smaller and less certain shift.
- Over the full historical record from 1879 to 2024, every season has been warming at a sustained long-run rate. All four seasons fall in the range of about 0.09 to 0.13 degrees Celsius per decade, with autumn warming fastest and summer warming slowest.
- The pace of warming strengthened after around 1975. Warming rates estimated over the most recent decades are consistently higher than those calculated over the full historical record in every season.
- A seasonal monitoring framework has been defined, assigning any incoming season a status of Reference, Watch, Action, or Escalation based on how far that season sits from the 1991 to 2020 reference distribution.

### Recommendation

Update all internal planning benchmarks to the 1991 to 2020 reference window. Assumptions built on the 1961 to 1990 baseline are now understating typical seasonal temperatures by roughly 0.6 to 1.1 degrees Celsius depending on the season, and this gap is likely to widen as the long-run warming trend continues. The main adjustment required is a revision of existing plans that use the older baseline, since any thresholds or targets set against 1961 to 1990 conditions will be exceeded more frequently than originally intended. Implement the four-level threshold framework from the next seasonal briefing cycle to enable consistent monitoring going forward.

### Evidence anchors

See Figure 1 in Section 3.2, Figure 2 in Section 4.1, Figure 3 in Section 4.2, Figure 4 in Section 4.3, and Table 1 in Section 4.4.

### Key uncertainties

The winter warming signal points in the same direction as the other three seasons, but its magnitude is less certain. The range of plausible values for the winter baseline shift includes zero, so winter conclusions should be treated with more caution than spring, summer, and autumn. All findings are specific to the Central England regional aggregate and do not describe variation at a finer local scale. The monitoring thresholds are designed for practical use in seasonal briefings, and different threshold settings would trigger alerts at different rates, which is worth bearing in mind if the framework is reviewed in future.

## Client brief and scope

### What the client asked

The brief requested a clear, evidence-based assessment of seasonal temperature change in Central England using the Met Office HadCET records. The practical goal was to support three linked planning decisions, covering whether a sustained shift in seasonal temperatures has taken place over recent decades, which seasons have changed the most and whether the change looks more like a gradual trend or a distinct step shift, and what metrics and thresholds to use for internal monitoring and communication.

### What this analysis covers

The analysis uses HadCET seasonal temperature records from 1879 to 2024, covering mean, minimum, and maximum temperatures across winter, spring, summer, and autumn. The primary baseline comparison is between a recent 30-year reference period (1991 to 2020) and an earlier historical period (1961 to 1990). Long-run warming rates are estimated over the full historical window. A shorter 10-year window covering 2015 to 2024 is also used as a near-term signal to detect whether the most recent decade sits above the 30-year reference distribution.

### What falls outside this scope

This report does not analyse sub-regional variation within Central England. It works with seasonal-level temperature summaries, so short-duration events such as individual heatwaves or cold spells at a daily or weekly scale are not captured. It does not make projections about future temperature levels.

## Data and context

### Where the data comes from

The analysis draws on three official seasonal temperature series from the Met Office HadCET archive (v2.1.0.0), covering seasonal mean, minimum, and maximum temperatures for the Central England region. Each series provides one value per season per year across a very long historical span. The mean series extends back to 1659, whilst the minimum and maximum series each begin in 1878. Because comparing all three temperature types requires overlapping records, all quantitative analysis in this report uses the shared period from 1879 to 2024.

### What the data shows

```{r eda-annotated-series-plot, fig.width=9, fig.height=3.5, fig.cap="Annual HadCET seasonal temperatures by season, with rolling 30-year means."}
p_eda_annotated <- make_eda_annotated_series_plot(save_plot = TRUE)
p_eda_annotated
```

\FloatBarrier

Figure 1 shows four panels, one for each season. In each panel, the faint dots represent the raw temperature recorded in each individual year, and the coloured lines trace the 30-year running average for mean, minimum, and maximum temperature. The two shaded bands mark the older reference window (1961 to 1990) and the more recent one (1991 to 2020). In every panel, the running average lines rise clearly from left to right, with the upward movement becoming steeper from the 1970s onward, and both shaded windows sit visibly above the temperature levels in the early part of the record. This consistent pattern across all four seasons establishes that the changes described in Section 4 are part of a broad, sustained process and are not the product of one or two unusual years.

### A note on the data

Because each data point summarises a full three-month season, shorter events within that period are not separately visible. A particularly hot fortnight in August, for instance, would contribute to the summer average but would not appear as an individual observation. A small number of historically exceptional seasons are present in the data, including winter 1963 and summer 1976. These were retained in the analysis, and later checks confirmed that they do not drive any of the headline results, so there is no need to treat them as special cases in the monitoring framework.
